---
title: "Gradient Descent"
author: "Kristi Angel"
date: "2/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# Programming Exercise 1: Linear Regression   
This exercise is intended to replicate the first assignment in Andrew Ng's Machine Learning Coursera course using R rather than Matlab.  

## Introduction  
In this exercise, you will implement linear regression and get to see it work on data. Before starting on this programming exercise, we strongly recommend watching the video lectures and completing the review questions for the associated topics.  

Files included in this exercise:  

* ex1.m - Octave/MATLAB script that steps you through the exercise  
* ex1_multi.m - Octave/MATLAB script for the later parts of the exercise  
* ex1data1.txt - Dataset for linear regression with one variable  
* ex1data2.txt - Dataset for linear regression with multiple variables  
* submit.m - Submission script that sends your solutions to our servers  
* [$\star$] warmUpExercise.m - Simple example function in Octave/MATLAB 
* [$\star$] plotData.m - Function to display the dataset  
* [$\star$] computeCost.m - Function to compute the cost of linear regression  
* [$\star$] gradientDescent.m - Function to run gradient descent  
* [$\dagger$] computeCostMulti.m - Cost function for multiple variables  
* [$\dagger$] gradientDescentMulti.m - Gradient descent for multiple variables   
* [$\dagger$] featureNormalize.m - Function to normalize features  
* [$\dagger$] normalEqn.m - Function to compute the normal equations  

[$\star$] indicates files you will need to complete   
[$\dagger$] indicates optional exercises  


```{r get_data}
population_profit <- read_csv("../data/ex1data1.txt", col_names = FALSE)
colnames(population_profit) <- c("population", "profit")
```

## Linear regression with one variable  
In this part of this exercise, you will implement linear regression with one variable to predict profits for a food truck. Suppose you are the CEO of a restaurant franchise and are considering different cities for opening a new outlet. The chain already has trucks in various cities and you have data for profits and populations from the cities.  

You would like to use this data to help you select which city to expand to next.  

The file ex1data1.txt contains the dataset for our linear regression problem. The first column is the population of a city and the second column is the profit of a food truck in that city. A negative value for profit indicates a loss.  

## Visualizing the Data  
Generating a scatterplot to better understand the data 

```{r scatter}
theme <- theme_classic() + 
            theme(plot.title = element_text(color = "gray30", size=18, face="bold"), 
                  axis.title.x = element_text(color = "gray34"),
                  axis.title.y = element_text(color = "gray34"))
population_profit %>%
  ggplot(aes(x = population, y = profit)) + 
  geom_point(color = "blue4", size = 5, alpha = 0.65) + 
  ggtitle("Restaurant Profits by City Population") + 
  ylab("Profit in $10,000's") + 
  xlab("Population of City in 10,000's") +
  theme
```

## Gradient Descent  
In this part, you will fit the linear regression parameters θ to our dataset using gradient descent.

The objective of linear regression is to minimize the cost function, 

$$J(\theta) = \frac{1}{2m}\sum_{i=1}^{m}(h_{\theta}(x^{(i)}) - y^{(i)})^2$$  

where the hypothesis hθ(x) is given by the linear model

$$h_{\theta}(x) = \theta^{T}x = \theta_0 + \theta_1x_1$$
Recall that the parameters of your model are the θj values. These are the values you will adjust to minimize cost J(θ). One way to do this is to use the batch gradient descent algorithm. In batch gradient descent, each iteration performs the update  

$$\theta_j := \theta_j - \alpha\frac{1}{m}\sum_{i=1}^{m}(h_{\theta}(x^{(i)}) - y^{(i)})x^{(i)}_{j}$$
With each step of gradient descent, your parameters θj come closer to the optimal values that will achieve the lowest cost J(θ).  

**Implementation Note:** We store each example as a row in the the X matrix in Octave/MATLAB. To take into account the intercept term (θ0), we add an additional first column to X and set it to all ones. This allows us to treat θ0 as simply another ‘feature’.  

```{r design_matrix}
# Add a column of ones to x
# initialize fitting parameters
x <- population_profit %>%
      select(population) %>%
      mutate(bias = rep(1, nrow(.))) %>%
      select(bias, population)
X <- as.matrix(x)
head(X)
y <- population_profit %>%
      select(profit)
y <- as.matrix(y)
theta <- as.matrix(rep(0, ncol(X)))
```

As you perform gradient descent to learn minimize the cost function J(θ), it is helpful to monitor the convergence by computing the cost. In this section, you will implement a function to calculate J(θ) so you can check the convergence of your gradient descent implementation.  

Your next task is to complete the code in the file computeCost.m, which is a function that computes J(θ). As you are doing this, remember that the variables X and y are not scalar values, but matrices whose rows represent the examples from the training set.  

Once you have completed the function, the next step in ex1.m will run computeCost once using θ initialized to zeros, and you will see the cost printed to the screen.  

In the first example, you should expect to see a cost of 32.07.  In the second you should expect to see an expected cost value of 54.24.

```{r cost_function}
costFunction <- function(X, theta, y){
  m <- nrow(y)
  predicted <- X %*% theta
  sum_square_err <- sum((predicted - y) ^ 2)
  j_theta <- sum_square_err / (2* m)
  return(j_theta)
}

costFunction(X, theta, y)

theta2 = matrix(c(-1, 2), nrow = 2)
costFunction(X, theta2, y)
```


```{r gradient_descent_algo}
iterations <- 1500
alpha <-  0.01
cost <- rep(NA, iterations)
  
gradientDescent <- function(X, y, alpha, iterations){
  theta <- as.matrix(rep(0, ncol(X)))
  m <- length(y)
  
  for(i in 1:iterations){
    predicted <- X %*% theta # 97x1 matrix
    theta <- theta - ((alpha / m) * (t(X) %*% (predicted - y)))
    cost[i] <- costFunction(X, theta, y)
  }
  return(list(theta, cost))
}

# might also want to define a convergence threshold and stopping criteria
model <- gradientDescent(X, y, alpha, iterations)
```

The theta values are: $\theta_0$ = `r round(model[[1]][1], 4)` and $\theta_1$ = `r round(model[[1]][2], 4)`. The cost is minimized at $J(\theta)$ = `r round(model[[2]][length(model[[2]])], 4)`.

We can observe how the cost function decreases with each iteration. 

```{r cost_v_iteration}
cost <- model[[2]]
iteration <- 1:1500
data <- tibble(iteration, cost)
data %>% ggplot(aes(x = iteration, y = cost)) +
  geom_point(color = "blue4", alpha = 0.075, size = 3) + 
  ggtitle("Cost Fuction As a Function of Algorithm Iteration") + 
  theme

```

### Fitting the Model
Let's see how well our parameters fit the data.

```{r model_fit}
y_pred <- model[[1]][1] + model[[1]][2] * population_profit$population
p <- population_profit %>% ggplot(aes(population, profit)) +
  geom_point(color = "blue4", alpha = 0.65, size = 5)
p + geom_line(aes(x = population_profit$population, y = y_pred), size = 1.5, color = "grey25") + 
  theme +
  ggtitle("Linear Regression: Profit as a Function of Population") + 
  xlab("Population in 10,000s") + 
  ylab("Profit in 10,000s")
```


## Linear regression with multiple variables  
In this part, you will implement linear regression with multiple variables to predict the prices of houses. Suppose you are selling your house and you want to know what a good market price would be. One way to do this is to first collect information on recent houses sold and make a model of housing prices.  

The file ex1data2.txt contains a training set of housing prices in Portland, Oregon. The first column is the size of the house (in square feet), the second column is the number of bedrooms, and the third column is the price of the house.  

The ex1 multi.m script has been set up to help you step through this exercise.


```{r}
home_prices <- read_csv("../data/ex1data2.txt", col_names = FALSE)
colnames(home_prices) <- c("sq_feet", "n_bedrooms", "price")
```

## Feature Normalization  
The ex1 multi.m script will start by loading and displaying some values from this dataset. By looking at the values, note that house sizes are about 1000 times the number of bedrooms. When features differ by orders of mag- nitude, first performing feature scaling can make gradient descent converge much more quickly.  

Your task here is to complete the code in featureNormalize.m to:  
• Subtract the mean value of each feature from the dataset.  
• After subtracting the mean, additionally scale (divide) the feature values by their respective “standard deviations.”  

The standard deviation is a way of measuring how much variation there is in the range of values of a particular feature (most data points will lie within ±2 standard deviations of the mean); this is an alternative to taking the range of values (max-min). In Octave/MATLAB, you can use the “std” function to compute the standard deviation. For example, inside featureNormalize.m, the quantity X(:,1) contains all the values of x1 (house sizes) in the training set, so std(X(:,1)) computes the standard deviation of the house sizes. At the time that featureNormalize.m is called, the extra column of 1’s corresponding to x0 = 1 has not yet been added to X (see ex1 multi.m for details).  

You will do this for all the features and your code should work with datasets of all sizes (any number of features / examples). Note that each column of the matrix X corresponds to one feature.  

**Implementation Note:** When normalizing the features, it is important to store the values used for normalization - the mean value and the stan- dard deviation used for the computations. After learning the parameters from the model, we often want to predict the prices of houses we have not seen before. Given a new x value (living room area and number of bed- rooms), we must first normalize x using the mean and standard deviation that we had previously computed from the training set.

```{r}

```

## Gradient Descent  
Previously, you implemented gradient descent on a univariate regression problem. The only difference now is that there is one more feature in the matrix X. The hypothesis function and the batch gradient descent update rule remain unchanged.  

You should complete the code in computeCostMulti.m and gradientDescentMulti.m to implement the cost function and gradient descent for linear regression with multiple variables. If your code in the previous part (single variable) already supports multiple variables, you can use it here too.  

Make sure your code supports any number of features and is well-vectorized. You can use ‘size(X, 2)’ to find out how many features are present in the dataset.  

**Implementation Note:** In the multivariate case, the cost function can also be written in the following vectorized form:  
# formula  
where  
# formula  
The vectorized version is efficient when you’re working with numerical computing tools like Octave/MATLAB. If you are an expert with matrix operations, you can prove to yourself that the two forms are equivalent.  

```{r}

```

## Optional (ungraded) exercise: Selecting learning rates  
In this part of the exercise, you will get to try out different learning rates for the dataset and find a learning rate that converges quickly. You can change the learning rate by modifying ex1 multi.m and changing the part of the code that sets the learning rate.  

The next phase in ex1 multi.m will call your gradientDescent.m func- tion and run gradient descent for about 50 iterations at the chosen learning rate. The function should also return the history of J(θ) values in a vector J. After the last iteration, the ex1 multi.m script plots the J values against the number of the iterations.  

If you picked a learning rate within a good range, your plot look similar Figure 4. If your graph looks very different, especially if your value of J(θ) increases or even blows up, adjust your learning rate and try again. We rec- ommend trying values of the learning rate α on a log-scale, at multiplicative steps of about 3 times the previous value (i.e., 0.3, 0.1, 0.03, 0.01 and so on). You may also want to adjust the number of iterations you are running if that will help you see the overall trend in the curve.  
```{r}

```

Notice the changes in the convergence curves as the learning rate changes. With a small learning rate, you should find that gradient descent takes a very long time to converge to the optimal value. Conversely, with a large learning rate, gradient descent might not converge or might even diverge!  

Using the best learning rate that you found, run the ex1 multi.m script to run gradient descent until convergence to find the final values of θ. Next, use this value of θ to predict the price of a house with 1650 square feet and 3 bedrooms. You will use value later to check your implementation of the normal equations. Don’t forget to normalize your features when you make this prediction!  

## Normal Equations  
In the lecture videos, you learned that the closed-form solution to linear regression is  
# formula  
Using this formula does not require any feature scaling, and you will get an exact solution in one calculation: there is no “loop until convergence” like in gradient descent.
Complete the code in normalEqn.m to use the formula above to calcu- late θ. Remember that while you don’t need to scale your features, we still need to add a column of 1’s to the X matrix to have an intercept term (θ0). The code in ex1.m will add the column of 1’s to X for you.  

```{r}

```

Optional (ungraded) exercise: Now, once you have found θ using this method, use it to make a price prediction for a 1650-square-foot house with 3 bedrooms. You should find that gives the same predicted price as the value you obtained using the model fit with gradient descent (in Section 3.2.1).  
